Type-Safe Modular Parsing

### 2017-01-07 paper/paper.tex
Polishing overview & trying a more concise example.
Revising and Splitting Overview
Added S3, inheritance-based approach.
Revised S4.
Revised S5. Related work to be revised.
Fix

% "define" code highlights for Java and Scala
\lstdefinelanguage{JavaScala}{
  morekeywords={public,int,interface,implements,default,
    abstract,case,catch,class,def,static,%
    do,else,extends,false,final,finally,%
    for,if,implicit,import,match,mixin,%
    new,null,object,override,package,%
    private,protected,requires,return,sealed,%
    super,this,throw,trait,true,try,%
    type,var,while,yield,with,val},
  otherkeywords={=>,<-,<\%,<:,>:,\#,@},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[n]{/*}{*/},
  morestring=[b]",
  morestring=[b]',
  morestring=[b]"""
}

@@

\input{sections/Overview.tex}

\input{sections/PackratParsers.tex}

\input{sections/OpenRecursion.tex}

\input{sections/ObjectAlgebras.tex}
\input{sections/ParsingLibrary.tex}

% Object Algebras
% type synonyms, Parser[Expr], Fix[..]
% more: two pretty-printers OldPretty and NewPretty, they are separated to show modularity.
% subsection: modify the code using List[E] and type bounds.
% more subsections: lexing, traits, operations, language composition (more extensibility)
% use paragraphs
%...

>>>>

% "define" code highlights for Java and Scala
\lstdefinelanguage{JavaScala}{
  morekeywords={public,int,interface,implements,default,
    abstract,case,catch,class,def,static,%
    do,else,extends,false,final,finally,%
    for,if,implicit,import,match,mixin,%
    new,null,object,override,package,%
    private,protected,requires,return,sealed,%
    super,this,throw,trait,true,try,%
    type,var,while,yield,with,val},
  otherkeywords={=>,<\%,<:,>:,\#,@},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[n]{/*}{*/},
  morestring=[b]",
  morestring=[b]',
  morestring=[b]"""
}

@@

\input{sections/PackratParsing.tex}

\input{sections/Inheritance.tex}

\input{sections/ObjectAlgebras.tex}

\input{sections/Discussion.tex}

### 2017-01-07 paper/sections/PackratParsing.tex
Fix

A left-recursive grammar for a simple arithmetic expression language can be written in a straightforward way. An alternative grammar, that eliminates left recursion is also presented on the right side:

\begin{tabular}{m{0.4\linewidth}m{0.4\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <expr> `+' <int>
\end{grammar}
&
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> <expr'>

<expr'> ::= <empty> \alt `+' <int> <expr'>
\end{grammar}
\end{tabular}

The main problem with left-recursion elimination is that it is a
\emph{global} transformation on a grammar. Given that the full grammar
is known, then it is possible to remove all left-recursive
cases. However, when doing modular parsing, the full grammar is not
known!

For the example above, if we extend the original grammar to support
subtraction, we must analyse the full grammar again to rewrite it.

\begin{tabular}{m{0.4\linewidth}m{0.4\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <expr> `+' <int> \alt <expr> `-' <int>
\end{grammar}
&
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> <expr'>

<expr'> ::= <empty> \alt `+' <int> <expr'> \alt `-' <int> <expr'>
\end{grammar}
\end{tabular}

Another issue of left-recursion elimination is that it requires extra
bookkeeping work to retain the original semantics. For example, the
expression $1+2-3$ is parsed as $(1+2)-3$ in the left-recursive
grammar, but after rewrite the result is $1((+2)-3))$. The parse tree
must be transformed to recover its structure.

\bruno{Huang, please fix this example ASAP.}
\bruno{This example does not appear to be problematic: adding ``-''
  does not affect the rest of the grammar. Isn't the problem that
  sometimes adding a new case to the grammar, requires changing other
parts? Otherwise I don't see why left-recursion is a problem, and we
cannot simply claim that it is a problem. You have to make a better
effort to find a problematic grammar.
}

>>>>

The grammar below represents a simple arithmetic language containing only integers. Its corresponding parser is shown on the right side, written in Parsec.

\begin{tabular}{m{0.4\linewidth}m{0.5\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int>
\end{grammar}
&
\begin{lstlisting}[language=PlainCode]
parseExpr = parseInt
\end{lstlisting}
\end{tabular}

Consider we extended the language by adding subtractions to the grammar. If we still write its parser by directly following the grammar structure, the new left-recursive case will get the parser into an infinite loop. Namely, \inlinecode{parseExpr} and \inlinecode{parseSub} call each other and never stop.

\begin{tabular}{m{0.4\linewidth}m{0.5\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <expr> `-' <int>
\end{grammar}
&
\begin{lstlisting}[language=PlainCode]
parseExpr = parseSub <|> parseInt
parseSub = do
  e <- parseExpr
  ...
\end{lstlisting}
\end{tabular}

To solve this issue, we have to rewrite the grammar as below to eliminate left-recursion and build the parser based on this new grammar.

\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> <expr'>

<expr'> ::= <empty> \alt `-' <int> <expr'>
\end{grammar}

After left-recursion elimination, the structure of grammar is changed, as well as its corresponding parser. In a modular setting, it is possible but unnecessorily compilcated to analyse the grammar and rewrite it when doing extension. Anticipating that every non-terminal has left-recursive rules, even though some do not, is helpful for extension. However, it is inconvenient and introduces extra complexity for representation of grammar and implementation of parser.

Another issue of left-recursion elimination is that it requires extra
bookkeeping work to retain the original semantics. For example, the
expression $1-2-3$ is parsed as $(1-2)-3$ in the left-recursive
grammar, but after rewrite the result is $1((-2)-3))$. The parse tree
must be transformed to recover its structure.
